Structured Versus Decorated Cospans

### 2020-11-13 cospans.tex
small edits

\begin{thm}\label{SC}
Given categories $\X$ and $\A$ with finite colimits and $L \maps \A \to \X$ a finitely cocontinuous functor, there exists a symmetric monoidal double category $_L\lCsp(\X)$ in which
\begin{itemize}
\item an object is an object of $\A$,
\item a vertical morphism is a morphism of $\A$,
\item a horizontal 1-cell $a\tickar b$ is a cospan in $\X$ of the form
\begin{equation}\label{eq:structuredcospan}
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$L(a)$};
\node (B) at (1,0) {$x$};
\node (C) at (2,0) {$L(b)$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$$} (B)
(C) edge node[above]{$$} (B);
\end{tikzpicture}
\end{equation}
\item a 2-morphism is a map of cospans of $\X$ of the form
\begin{equation}\label{eq:2cellsStrCsp}
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$L(a)$};
\node (B) at (1,0) {$x$};
\node (C) at (2,0) {$L(b)$};
\node (A') at (0,-1) {$L(a')$};
\node (B') at (1,-1) {$x'$};
\node (C') at (2,-1) {$L(b')$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$$} (B)
(C) edge node[above]{$$} (B)
(A') edge node[above]{$$} (B')
(C') edge node[above]{$$} (B')
(A) edge node [left]{$L(f)$} (A')
(B) edge node [left]{$\alpha$} (B')
(C) edge node [right]{$L(g)$} (C');
\end{tikzpicture}
\end{equation}
\end{itemize}
The symmetric monoidal structure is formed by using binary coproducts and the initial object.
\end{thm}

@@

The desired equivalence between these two different double categorical network frameworks, namely structured cospans and decorated cospans,
is established by the following theorem. We denote by $\bicat{Rex}$ the 2-category of finitely cocomplete categories and finitely cocontinuous functors.

\begin{thm}\label{Equiv}
Suppose $\ca{A}$ is a finitely cocomplete category and $(F,\phi,\phi_0)\maps(\ca{A},+,0) \to (\bicat{Cat},\times,\mathbf{1})$ is a symmetric lax monoidal pseudofunctor that factors through $\bicat{Rex}$ as an ordinary pseudofunctor. 
Then the symmetric monoidal double category $F\lCsp$ of decorated cospans given in \cref{thm:decoratedcospans} is equivalent to the symmetric monoidal double category $\scsp{L_F}{\inta F}$ of structured cospans given in \cref{SC}, where $L_F \maps \A \to \inta{F}$ is the left adjoint of the induced Grothendieck opfibration $\U_F\maps \inta{F} \to \A$. %by \cref{thm:Grothendieck}, 
%under the assumptions of \cref{cor:fcocMonGroth}.
\end{thm}

The conditions of the above theorem relate to the existence of left adjoints for opfibrations; we first sketch this underlying framework, and then we proceed to the proof of the theorem.

The basics of fibration theory needed for our purposes are recalled in \cref{sec:fibrations}. 
In \cite{MV}, the classic Grothendieck construction is generalized to the monoidal setting: lax monoidal structures on a pseudofunctor $F\maps\A\to\bicat{Cat}$ bijectively correspond to monoidal structures on the total category $\inta F$ such that the induced opfibration is a strict monoidal functor and $\otimes_{\inta F}$ preserves liftings, see \cite[Thm. 3.10]{MV}. If the monoidal category $\A$ is in fact cocartesian, there is a further correspondence of the two structures with an ordinary pseudofunctor $F\maps\A\to\bicat{MonCat}$ into the 2-category of monoidal categories, strong monoidal functors and monoidal natural transformations, summarized as
\begin{gather}
\textrm{lax monoidal pseudofunctors }F\maps(\A,+,0)\to(\bicat{Cat},\times) \notag\\
\Updownarrow \notag\\
\textrm{monoidal opfibrations }U\maps(\X,\otimes)\to(\A,+) \label{monGroth}\\
\Updownarrow \notag\\
\textrm{pseudofunctors }F\maps\A\to\bicat{MonCat} \notag
\end{gather}
The second equivalence was in fact earlier observed by Shulman in \cite{Shul2}. In more detail, if $(\phi,\phi_0)$ is the lax monoidal structure of $F$, each fibre category $\X_a=F(a)$ obtains a monoidal structure via
\begin{gather}\label{eq:explicitstructure1}
\otimes_a\maps F(a)\times F(a)\xrightarrow{\phi_{a,a}}F(a+a)\xrightarrow{F(\nabla)}F(a)\\
I_x\maps\mathbf{1}\xrightarrow{\phi_0}F(0)\xrightarrow{F(!)}F(a)\nonumber
\end{gather}
Moreover, these correspondences further restrict to the case when the Grothendieck category is specifically cocartesian monoidal itself; for more details, see \cite[Cor. 4.8]{MV}. In that case, the monoidal opfibration clauses for $U\maps(\X,+)\to(\A,+)$ translate to the functor (strictly) preserving coproducts and the initial object, and bijectively correspond to pseudofunctors $F\maps\A\to\Rex $ into the 2-category of esian categories, functors that preserve coproducts and initial objects and natural transformations.

@@

A \define{lax monoidal} pseudofunctor between monoidal bicategories $F\maps\bicat{A}\to\bicat{B}$ is a pseudofunctor equipped with pseudonatural transformations with components $\mu_{a,b}\maps Fa\otimes Fb\to F(a\otimes b)$ and $\mu_0\maps I\to FI$ along with coherent invertible modifications for associativity and unitality. This is commonly also called `weak monoida' pseudofunctor. A \define{symmetric lax monoidal} pseudofunctor between symmetric monoidal bicategories comes with invertible $F(b)\circ \mu_{a,b}\cong \mu_{b,a}\circ b$, where $b$ is the braiding. 
%!!!Christina: Hiding various details under the rug - e.g. what a symmetric monoidal bicategory is. Should be fine? Also, not sure if we need monoidal transformations?

@@

\bibitem{MV} J.\ Moeller and C.\ Vasilakopoulou, Monoidal Grothendieck construction. Available as \href{https://arxiv.org/abs/1809.00727}{arXiv:1809.00727}.

>>>>

\begin{thm}\label{SC}
Given categories $\X$ and $\A$ with finite colimits and $L \maps \A \to \X$ a finitely cocontinuous functor, there exists a symmetric monoidal double category $_L\lCsp(\X)$ in which
\begin{itemize}
\item an object is an object of $\A$,
\item a vertical morphism is a morphism of $\A$,
\item a horizontal 1-cell $a\tickar b$ is a cospan in $\X$ of the form
\begin{equation}\label{eq:structuredcospan}
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$L(a)$};
\node (B) at (1,0) {$x$};
\node (C) at (2,0) {$L(b)$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$$} (B)
(C) edge node[above]{$$} (B);
\end{tikzpicture}
\end{equation}
\item a 2-morphism is a map of cospans in $\X$ of the form
\begin{equation}\label{eq:2cellsStrCsp}
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$L(a)$};
\node (B) at (1,0) {$x$};
\node (C) at (2,0) {$L(b)$};
\node (A') at (0,-1) {$L(a')$};
\node (B') at (1,-1) {$x'$};
\node (C') at (2,-1) {$L(b')$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$$} (B)
(C) edge node[above]{$$} (B)
(A') edge node[above]{$$} (B')
(C') edge node[above]{$$} (B')
(A) edge node [left]{$L(f)$} (A')
(B) edge node [left]{$\alpha$} (B')
(C) edge node [right]{$L(g)$} (C');
\end{tikzpicture}
\end{equation}
\end{itemize}
The symmetric monoidal structure is formed by using binary coproducts and the initial object.
\end{thm}

@@

The following theorem establishes the desired equivalence between these two different double categorical network frameworks of structured and decorated cospans in a formal way. We denote by $\bicat{Rex}$ the 2-category of finitely cocomplete categories and finitely cocontinuous functors, and the symmetric monoidal double structure of decorated cospans is worked out in \cref{thm:decoratedcospans,DC}.

\begin{thm}\label{Equiv}
Suppose $\ca{A}$ has finite colimits and $(F,\phi,\phi_0)\maps(\ca{A},+,0) \to (\bicat{Cat},\times,\mathbf{1})$ is a symmetric lax monoidal pseudofunctor that factors through $\bicat{Rex}$ as an ordinary pseudofunctor. 
Then the symmetric monoidal double category $F\lCsp$ of decorated cospans is equivalent to the symmetric monoidal double category $_L\lCsp(\inta F)$ of structured cospans, where $L \maps \A \to \inta F$ is the left adjoint of the induced Grothendieck opfibration $\U\maps \inta F \to \A$.
\end{thm}

The conditions of the above theorem relate to the existence of left adjoints for opfibrations; we first sketch this underlying framework, and then we proceed to the proof of the theorem. The basics of fibration theory needed for our purposes are recalled in \cref{sec:fibrations}.

In \cite{MV}, the classical Grothendieck construction is generalized to the monoidal setting: lax monoidal structures on a pseudofunctor $F\maps\A\to\bicat{Cat}$ bijectively correspond to monoidal structures on the total category $\inta F$ such that the induced opfibration is a strict monoidal functor and $\otimes_{\inta F}$ preserves cocartesian liftings, see \cite[Thm.~3.13]{MV}. If the monoidal category $\A$ is in fact cocartesian, there is a further correspondence of the two structures with an ordinary pseudofunctor $F\maps\A\to\bicat{MonCat}$ into the 2-category of monoidal categories, strong monoidal functors and monoidal natural transformations, summarized as
\begin{gather}
\textrm{lax monoidal pseudofunctors }F\maps(\A,+,0)\to(\bicat{Cat},\times) \notag\\
\Updownarrow \notag\\
\textrm{monoidal opfibrations }U\maps(\X,\otimes)\to(\A,+) \label{monGroth}\\
\Updownarrow \notag\\
\textrm{pseudofunctors }F\maps\A\to\bicat{MonCat} \notag
\end{gather}
The second equivalence was in fact earlier observed by Shulman in \cite{Shul2}. In more detail, if $(\phi,\phi_0)$ is the lax monoidal structure of $F$, each fibre category $\X_a=F(a)$ obtains a monoidal structure via
\begin{gather}\label{eq:explicitstructure1}
\otimes_a\maps F(a)\times F(a)\xrightarrow{\phi_{a,a}}F(a+a)\xrightarrow{F(\nabla)}F(a)\\
I_x\maps\mathbf{1}\xrightarrow{\phi_0}F(0)\xrightarrow{F(!)}F(a)\nonumber
\end{gather}
Moreover, these correspondences further restrict to the case when the Grothendieck category is specifically cocartesian monoidal itself; for more details, see \cite[Cor. 4.8]{MV}. In that case, the monoidal opfibration clauses for $U\maps(\X,+)\to(\A,+)$ translate to the functor (strictly) preserving coproducts and the initial object, and bijectively correspond to pseudofunctors $F\maps\A\to\Rex $ into the 2-category of esian categories, functors that preserve coproducts and initial objects and natural transformations.

@@

A \define{lax monoidal} pseudofunctor between monoidal bicategories $F\maps\bicat{A}\to\bicat{B}$ is a pseudofunctor equipped with pseudonatural transformations with components $\mu_{a,b}\maps Fa\otimes Fb\to F(a\otimes b)$ and $\mu_0\maps I\to FI$ along with coherent invertible modifications for associativity and unitality. This is commonly also called `weak monoidal' pseudofunctor. A \define{symmetric lax monoidal} pseudofunctor between symmetric monoidal bicategories comes with invertible $F(b)\circ \mu_{a,b}\cong \mu_{b,a}\circ b$, where $b$ is the braiding. 
%!!!Christina: Hiding various details under the rug - e.g. what a symmetric monoidal bicategory is. Should be fine? Also, not sure if we need monoidal transformations?

@@

\bibitem{MV} J.\ Moeller and C.\ Vasilakopoulou, Monoidal Grothendieck construction, \emph{Theory Appl.\ Categ.} \textbf{35} (2020), no. 31, 1159--1207. Available as \href{https://arxiv.org/abs/1809.00727}{arXiv:1809.00727}.
